---
title: 'Triage Against the Machine: Can AI Reason Deliberatively? Part II'
author: "Francesco Veri, Gustavo Umbelino"
date: "`r Sys.Date()`"
bibliography: ../bibliography/refs.bib
link-citations: true
csl: ../bibliography/apsa.csl
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(readr)
library(readxl)
library(dplyr)
library(rstatix)

LLM_DATA_DIR <- "../llm_data"
SURVEY_FILE <- "../data/surveys_v5.xlsx"
EXEC_LOG_FILE <- paste(LLM_DATA_DIR, "exec_log.csv", sep = "/")
LLMS_FILE <- "../private/pt2_llms.csv"
CASES_FILE <- "../data/deliberative_cases.csv"
PROMPTS_FILE <- "../prompts/prompts.csv"
OUTPUT_DIR <- "pt2/data"

CONSISTENCY_RESULTS_FILE <- "data/consistency_results.csv"


IT <- 5

```

# Large-Language Models (LLMs) Preview

```{r models, warning=FALSE}

# get models info
models <- read_csv(LLMS_FILE, show_col_types = FALSE) %>%
  filter(included) # only included "included" models

models %>%
  select(provider, model) %>% 
  arrange(provider, model) %>%
  knitr::kable(caption = "LLMs", row.names = TRUE, col.names = c("Provider", "Model"))

```

Building on our previous analysis, we selected only top models.

# Cases

```{r cases, warning=FALSE}

cases <- read_csv(CASES_FILE, show_col_types = FALSE)

cases %>% arrange(survey) %>%
  knitr::kable(caption = "Deliberative Cases", row.names = TRUE)
```

# Surveys

```{r surveys, warning=FALSE}

# get deliberative survey names
survey_names <- unique(cases$survey)

get_surveys <- function(survey_names) {
  
  surveys <- list()
  
  # Iterate over each sheet in the workbook
  for (survey_name in survey_names) {

    # Read the current sheet into a data frame
    df <- read_excel(SURVEY_FILE, sheet = survey_name)
    
    # Check if required columns exist
    required_columns <- c("considerations", "policies", "scale_max", "q-method")
    missing_cols <- setdiff(required_columns, colnames(df))
    if (length(missing_cols) > 0) {
      cat(
        "Sheet",
        survey_name,
        "is missing the following columns:",
        paste(missing_cols, collapse = ", "),
        "\n\n"
      )
      next
    }
    
    # Calculate the number of non-NA rows in "considerations" column
    n_c <- sum(!is.na(df$considerations))
    
    # Calculate the number of non-NA rows in "policies" column
    n_p <- sum(!is.na(df$policies))
    
    # Extract integer values from "scale_max" column, assuming they are already integers
    scale_max <- as.integer(na.omit(df$scale_max))
    
    # Extract logical (boolean) values from "q-method" column
    q_method <- as.logical(na.omit(df$`q-method`))
    
    surveys[[length(surveys) + 1]] <- tibble(
      survey = survey_name,
      considerations = n_c,
      policies = n_p,
      scale_max,
      q_method,
    )
    
  }
  
  surveys <- bind_rows(surveys)
  surveys
  
}

surveys <- get_surveys(survey_names)

surveys %>% arrange(survey) %>%
  knitr::kable(caption = "Surveys", row.names = TRUE)

```

# Roles (System Prompts)

```{r prompts, warning=FALSE}

prompts <- read_csv(PROMPTS_FILE, show_col_types = FALSE)

prompts %>% 
  group_by(type) %>%
  summarise(N = n()) %>%
  knitr::kable(caption = "Number of Prompts by Type")

prompts %>% arrange(type) %>%
  select(-article) %>%
  knitr::kable(caption = "System Prompts", row.names = TRUE)

```

# LLM Data Collection

```{r format LLM data, include=FALSE}

# source("get_llm_data.R")

llm_data <- read_csv("data/llm_data_clean.csv", show_col_types = FALSE)

progress <- llm_data %>%
  group_by(model, survey, prompt_uid) %>%
  summarise(N = n()) %>%
  filter(model %in% models$model)

for (uid in unique(prompts$uid)) {
  
  for (survey in surveys$survey) {
    
    for (model in models$model) {
      
      # if model/survey/uid doesn't exist,
      n <- progress %>% filter(prompt_uid == uid, survey == !!survey, model == !!model) %>% nrow()
      if (n == 0) {
        new_row <- tibble(
          model,
          survey,
          prompt_uid = uid,
          N = 0
        )
        progress <- bind_rows(progress, new_row)
      }
      
    }
    
  }
  
}

progress %>%
  mutate(done = N >= IT) %>%
  arrange(model, survey, prompt_uid) %>%
  filter(!done) %>%
  knitr::kable(caption = "Data Generation Progress")

progress %>%
  group_by(model) %>%
  summarise(done = sum(N)) %>%
  mutate(total = nrow(surveys) * nrow(prompts) * IT,
         perc = done * 100 / total) %>%
  arrange(perc) %>%
  knitr::kable(caption = "Progress per Model", digits = 2)


progress %>%
  group_by(prompt_uid) %>%
  summarise(done = sum(N)) %>%
  mutate(total = nrow(surveys) * nrow(models) * IT,
         perc = done * 100 / total) %>%
  arrange(perc) %>%
  knitr::kable(caption = "Progress per Prompt", digits = 2)
  

```

# Internal Consistency
```{r}

consistency_results <- read_csv(CONSISTENCY_RESULTS_FILE, show_col_types = FALSE)

# Aggregate consistency_results by model and calculate summary statistics
consistency_results %>%
  group_by(provider, model) %>%
  select(!N) %>%
  get_summary_stats(type="common") %>%
  knitr::kable(caption = "Cronbach's Alpha Summary Across 8 Surveys and 21 prompts", digits = 2)

consistency_results %>%
  arrange(-alpha_all) %>%
  knitr::kable(caption = "Cronbach's Alpha Full Results", digits = 2)

```

# Plots Surveys vs. Prompt
```{r}

library(deliberr)

res <- list()

for (uid in unique(prompts$uid)) {
  
  for (survey in unique(surveys$survey)) {
    
    df <- llm_data %>%
      filter(survey == !!survey) %>%
      filter(prompt_uid == uid) %>%
      mutate(pnum = row_number()) # treat each response as a participant
  
    ic <- get_dri_ic(df)
    dri <- get_dri(ic)
    
    res[[length(res)+1]] <- tibble(
      prompt = uid,
      survey,
      dri
    )
    
    # plot(plot_dri_ic(ic, title = paste(survey,uid, sep = "/"), suffix = "All LLMs", DRI = dri))
    
  }
  
}

res <- bind_rows(res)

res %>%
  arrange(-dri) %>%
  knitr::kable(caption = "Prompt vs. Survey DRI across all LLMs", digits = 3)

```



# Testing deliberation package
```{r}

# install.packages("gumbelino/deliberr")

lsf.str("package:deliberr")

df <- llm_data %>%
  filter(survey == "ccps") %>%
  filter(prompt_uid %in% c("eld", "csk") ) %>%
  mutate(pnum = row_number()) # treat each response as a participant

ic <- get_dri_ic(df)
dri <- get_dri(ic)

# library(grid)
# library(ggplot2)

plot_dri_ic(ic, "Ecologists vs. Climate Skeptics", "LLMs", dri)


```


```{r}

library(psych)
get_dri_alpha <- function(data) {
      
      # Calculate Cronbach's Alpha for considerations
      considerations_data <- data %>%
        select(matches("^C\\d+$") & where(~!all(is.na(.))))
      
      if (nrow(considerations_data) > 1) {
        
        # Check if policies are all equal (no variance)
        # this can happen when there are few iterations
        c_all_equal <- all(apply(considerations_data, 1, function(row)
          all(row == considerations_data[1, ], na.rm = TRUE)), na.rm = TRUE)
        
        # NOTE: assign alpha = 1, which should NOT exist!
        if (c_all_equal) {
          alpha_considerations <- 1
        } else {
        alpha_considerations <- psych::alpha(
          considerations_data,
          check.keys = TRUE,
          warnings = FALSE,
        )$total$raw_alpha
        }
      } else {
        alpha_considerations <- NA
      }
      
      # Calculate Cronbach's Alpha for policies
      policies_data <- data %>%
        select(matches("^P\\d+$") & where(~!all(is.na(.))))
      
      if (nrow(policies_data) > 1) {
        
        # Check if policies are all equal (no variance)
        # this can happen when there are few iterations
        p_all_equal <- all(apply(policies_data, 1, function(row)
          all(row == policies_data[1, ], na.rm = TRUE)), na.rm = TRUE)
        
        # NOTE: assign alpha = 1, which should NOT exist!
        if (p_all_equal) {
          alpha_policies <- 1
        }
        
        # normal case, calculate alpha
        else {
        alpha_policies <- psych::alpha(
          policies_data,
          check.keys = TRUE,
          warnings = FALSE,
        )$total$raw_alpha
        }
      } else {
        alpha_policies <- NA
      }
      
      if (nrow(policies_data) > 1 && nrow(considerations_data) > 1) {
        all_data <- cbind(considerations_data, policies_data)
        alpha_all <- psych::alpha(
          all_data,
          check.keys = TRUE,
          warnings = FALSE,
        )$total$raw_alpha
      } else {
        alpha_all <- NA
      }
      
      # Store the results in the list
      result <- tibble(
        alpha_c = alpha_considerations,
        alpha_p = alpha_policies,
        alpha_all = alpha_all,
      )
      
      return(result)
  
}

```




## Climate analysis
```{r climate, include=FALSE}
# get cases
cc_cases <- cases %>% filter(topic == "climate")

# get surveys
cc_surveys <- unique(cc_cases$survey)

# get roles: ecologist ideology, climate perspectives, or devil's advocate
cc_roles <- prompts %>% filter(type != "ideology" | uid == "eco")

library(ggplot2)

res <- list()

for (model in models$model) {
  
  for (survey in cc_surveys) {
    
    data <- llm_data %>% filter(model == !!model, survey == !!survey) %>%
        mutate(pnum = row_number()) # treat each response as a participant
    ic <- get_dri_ic(data)
    dri <- get_dri(ic)
    alpha <- get_dri_alpha(data)
    
    title <- paste(model, survey, sep = "/")
    suffix <- paste("all", nrow(cc_roles), "climate roles")
    # plot <- plot_dri_ic(ic, title, suffix, DRI = dri)
    
    file_name <- paste0(model, "-", survey, ".png")
    
    # ggsave(
    #   paste("plots", file_name, sep="/"),
    #   plot,
    #   width = 8,
    #   height = 6
    # )
    
    res[[length(res)+1]] <- tibble(
      model,
      survey,
      role = "all",
      dri,
      alpha,
      n = nrow(data),
    )
    
    
    for (role in cc_roles$uid) {
      
      data <- llm_data %>%
        filter(model == !!model, survey == !!survey, prompt_uid == role) %>%
        mutate(pnum = row_number()) # treat each response as a participant
      ic <- get_dri_ic(data)
      dri <- get_dri(ic) 
      alpha <- get_dri_alpha(data)
      
      res[[length(res)+1]] <- tibble(
        model,
        survey,
        role,
        dri,
        alpha,
        n = nrow(data),
      )
      
    }
    
    
    
  }
  
}

res <- bind_rows(res)

# res %>%
#   arrange(model, survey, role) %>%
#   knitr::kable(caption = "DRI consistency cross 12 climate roles", digits = 3)

library(tidyr)
```



# Summary DRI
```{r dri-summary}
model_role_summary <- res %>%
  group_by(model, role) %>%
  summarise(mean_dri = mean(dri)) %>%
  pivot_wider(names_from = model, values_from = mean_dri)


# find best model for each row
model_role_summary$best <- 
  colnames(model_role_summary)[max.col(model_role_summary[, -1],
                                       ties.method = "first") + 1]

model_role_summary %>%
  knitr::kable(caption = "Mean DRI across models and roles", digits = 3, row.names = TRUE)


```


# Summary Cronbach's Alpha (Policies)
```{r}
model_role_summary <- res %>%
  group_by(model, role) %>%
  summarise(mean_alpha_p = mean(alpha_p)) %>%
  pivot_wider(names_from = model, values_from = mean_alpha_p)


# find best model for each row
model_role_summary$best <- 
  colnames(model_role_summary)[max.col(model_role_summary[, -1],
                                       ties.method = "first") + 1]

model_role_summary %>%
  knitr::kable(caption = "Mean alpha (policies) across models and roles", digits = 3, row.names = TRUE)


```


# Summary Cronbach's Alpha (Consideration)
```{r}
model_role_summary <- res %>%
  group_by(model, role) %>%
  summarise(mean_alpha_c = mean(alpha_c)) %>%
  pivot_wider(names_from = model, values_from = mean_alpha_c)


# find best model for each row
model_role_summary$best <- 
  colnames(model_role_summary)[max.col(model_role_summary[, -1],
                                       ties.method = "first") + 1]

model_role_summary %>%
  knitr::kable(caption = "Mean alpha (considerations) across models and roles", digits = 3, row.names = TRUE)


```



## Testing DRI formulas with human data
```{r}

human_data <- read_csv("../data/human_data.csv", show_col_types = FALSE)
test <- list()
for (case in unique(human_data$case)) {
  data <- human_data %>% filter(case == !!case, stage_id == 1)
  ic <- get_dri_ic(data)
  dri <- get_dri(ic, adjusted = FALSE)
  dri_adj <- get_dri(ic)
  
  cat(case, "|", dri, "|", dri_adj, "\n")
  
  test[[length(test)+1]] <- tibble(
    case,
    dri,
    dri_adj
  )
}

test <- bind_rows(test)

```


# References
