---
title: "Benchmarking AI's Deliberative Reasoning: Evaluating LLMs Against Human Collective Wisdo"
author: "Francesco Veri, Gustavo Umbelino"
date: "`r Sys.Date()`"
bibliography: bibliography/refs.bib
link-citations: true
csl: bibliography/apsa.csl
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(reticulate)
library(tidyverse)
library(scales)
library(readxl)
library(psych)
library(boot)
library(car)
library(ggpubr)
library(rstatix)
library(data.table)



OUTPUT_DIR <- "analysis"

LLMS_FILE <- "private/llms_v3.csv"
LLM_DATA_FILE <- "llm_data.csv"
DRI_FILE <- "DRIInd_LLMs_iterations.csv"

MAX_ITERATIONS <- 5

```


```{r helper functions}

read_output_csv <- function(file_name) {
  read_csv(paste(OUTPUT_DIR, file_name, sep = "/"), show_col_types = FALSE)
}

write_output_csv <- function(data, file_name) {
  write_csv(data, paste(OUTPUT_DIR, file_name, sep = "/"))
}

now_utc <- function() {
   now <- Sys.time()
   attr(now, "tzone") <- "UTC"
   now
}

remove_NA_cols <- function(data) {
   data[, colSums(is.na(data)) != nrow(data)]
}

```

# Get Models for Analysis
```{r models, warning=FALSE}

# get models info
models <- read_csv(LLMS_FILE, show_col_types = FALSE) %>%
  filter(included)

models %>%
  select(provider, model, is_reasoner) %>% 
  arrange(provider, model) %>%
  knitr::kable(caption = "LLMs", row.names = TRUE, col.names = c("Provider", "Model", "Is Reasoner?"))

```

## Get average DRI score for each model

Calculate average DRI for the first `r MAX_ITERATIONS` iterations for each model, case

```{r dri}

dri_data <- read_output_csv(DRI_FILE)

dri_data <- dri_data %>%
  filter(iteration <= MAX_ITERATIONS) %>%
  group_by(provider, model, survey, case) %>%
  summarise(
    N = n(),
    DRIPostV3_mean = mean(DRIPostV3, na.rm=TRUE),
    .groups = "drop"
  )

```


Get LLM data to calculate alpha
```{r model info, results = 'asis', warning=FALSE}

llm_data <- read_output_csv(LLM_DATA_FILE)

# only use included models for analysis
llm_data <- llm_data %>% 
  filter(
    model %in% dri_data$model,
    survey %in% dri_data$survey,
  )

```


## Calculate Alpha
```{r error=FALSE, warning=FALSE, message=FALSE}

# Initialize an empty list to store the alpha results
alpha_results <- list()

models_with_data <- llm_data %>%
  distinct(provider, model)

# if (output.file.exists(ALPHA_RESULTS_FILE)) {
#   last_alpha_results <- read_output_csv(ALPHA_RESULTS_FILE)
# }

# Iterate over each unique provider/model combination
for (row in 1:nrow(models_with_data)) {
  provider <- models_with_data[row, ]$provider
  model <- models_with_data[row, ]$model
  
  # filter the data for the current provider/model
  provider_model_data <- llm_data %>%
    filter(model == !!model)
  
  # iterate over each survey
  for (survey_name in unique(provider_model_data$survey)) {
    
    # filter the data for the current survey
    survey_data <- provider_model_data %>%
      filter(survey == !!survey_name) %>%
      
      # get only first x iterations, where x = MAX_ITERATIONS
      arrange(created_at) %>%
      head(MAX_ITERATIONS)
    
    # SKIP THIS ITERATION IF PREVIOUS RECORD EXISTS
    # get date of the most data generation
    # last_updated <- max(survey_data$created_at, na.rm = TRUE)
    # 
    # # if there is a previous record
    # if (exists("last_alpha_results")) {
    #   last_alpha <- last_alpha_results %>% filter(model == !!model, survey == !!survey_name)
    #   
    #   ## and last record is still valid, save it and skip
    #   ## valid records are more recent than the latest data generated
    #   if (nrow(last_alpha) == 1 && (last_alpha$created_at > last_updated)) {
    #     alpha_results[[length(alpha_results) + 1]] <- last_alpha
    #     next
    #   } else {
    #     cat("updating:", model, "/", survey_name, "\n")
    #   }
    # }
    
    # Calculate Cronbach's Alpha for considerations (C1..C50)
    considerations_data <- survey_data %>% select(C1:C50)
    
    if (nrow(considerations_data) > 1) {
      
      # Check if policies are all equal (no variance)
      # this can happen when there are few iterations
      c_all_equal <- all(apply(considerations_data, 1, function(row)
        all(row == considerations_data[1, ], na.rm = TRUE)), na.rm = TRUE)
      
      # TODO: FIXME! 
      # NOTE: assign alpha = 1, which should NOT exist!
      # if (c_all_equal) {
      #   alpha_considerations <- 1
      # } else {
        alpha_considerations <- psych::alpha(
          considerations_data,
          check.keys = TRUE,
          warnings = FALSE,
        )$total$raw_alpha
      # }
    } else {
      alpha_considerations <- NA
    }
    
    # Calculate Cronbach's Alpha for policies (P1..P10)
    policies_data <- survey_data %>% select(P1:P10)
    
    if (nrow(policies_data) > 1) {
      
      # Check if policies are all equal (no variance)
      # this can happen when there are few iterations
      p_all_equal <- all(apply(policies_data, 1, function(row)
        all(row == policies_data[1, ], na.rm = TRUE)), na.rm = TRUE)
      
      # NOTE: assign alpha = 1, which should NOT exist!
      if (p_all_equal) {
        alpha_policies <- 1
      }

      # normal case, calculate alpha
      else {
        alpha_policies <- psych::alpha(
          policies_data,
          check.keys = TRUE,
          warnings = FALSE,
        )$total$raw_alpha
      }
    } else {
      alpha_policies <- NA
    }
    
    if (nrow(policies_data) > 1 && nrow(considerations_data) > 1) {
      all_data <- cbind(considerations_data, policies_data)
      alpha_all <- psych::alpha(
        all_data,
        check.keys = TRUE,
        warnings = FALSE,
      )$total$raw_alpha
    } else {
      alpha_all <- NA
    }
    
    
    # Store the results in the list
    alpha_results[[length(alpha_results) + 1]] <- tibble(
      provider = provider,
      model = model,
      survey = survey_name,
      N = nrow(considerations_data),
      created_at = now_utc(),
      alpha_considerations = alpha_considerations,
      alpha_policies = alpha_policies,
      alpha_all = alpha_all
    )
  }
}

# Combine all results into a single data frame
alpha_results <- bind_rows(alpha_results)

rm(models_with_data)
rm(considerations_data)
rm(survey_data)
rm(policies_data)
rm(provider_model_data)


```


```{r join data}

data <- full_join(dri_data, alpha_results, by=join_by(provider, model,survey))

data <- data %>%
  filter(!model %like% "t=1")

```

# Analysis

## Alpha vs. DRI correlation
We want to plot the corelation between ALPHA and DRI across models and cases

### Normality tests

```{r}

test_var <- data$DRIPostV3_mean

ggdensity(test_var, 
          main = "Density plot of DRI",
          xlab = "DRI")

ggqqplot(test_var)

shapiro.test(test_var)

test_var <- data$alpha_policies

ggdensity(test_var, 
          main = "Density plot of Alpha (Policies)",
          xlab = "DRI")

ggqqplot(test_var)

shapiro.test(test_var)


```

We use Spearman correlation because both variables are not normally distributed.


```{r}

data %>%
ggscatter(x = "DRIPostV3_mean", y = "alpha_policies", 
          add = "reg.line", conf.int = TRUE, 
          #facet.by = "provider",
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "DRI", ylab = "Cronbach's Alpha (Policies)")




```

### Group data by Model

We grouped data by model. 
```{r}

grouped_data <- data %>%
  group_by(provider, model) %>%
  summarize(
    DRI = mean(DRIPostV3_mean, na.rm = TRUE),
    alpha = mean(alpha_policies, na.rm = TRUE)
  )

```

We tested normality again.

```{r}

test_var <- grouped_data$DRI

ggdensity(test_var, 
          main = "Density plot of DRI",
          xlab = "DRI")

ggqqplot(test_var)

shapiro.test(test_var)

test_var <- grouped_data$alpha

ggdensity(test_var, 
          main = "Density plot of Alpha (Policies)",
          xlab = "DRI")

ggqqplot(test_var)

shapiro.test(test_var)

```
And used Spearman correlation.

```{r}

labels <- grouped_data[which(grouped_data$alpha > 0.80 | grouped_data$DRI > 0.35), ]$model

nrow(grouped_data)

grouped_data %>%
  ggscatter(
    x = "DRI",
    y = "alpha",
    add = "reg.line",
    # ylim=c(0.7, 0.9),
    #xlim=c(-0.1, 0.5),
    #label = "model",
    #repel = TRUE,
    #label.select = labels,
    # label.rectangle = TRUE,
    conf.int = TRUE,
    cor.coef = TRUE,
    cor.method = "spearman",
    #title = "Spearman Correlation Grouped by Model",
    xlab = "LLM DRI (post-deliberation, individual-level)",
    ylab = "Cronbach's alpha (policies)"
  ) -> plot

plot

ggsave(
  paste(OUTPUT_DIR, "plots", "book1.png", sep = "/"),
  plot,
  width = 8,
  height = 4
)

cor.test(grouped_data$DRI, grouped_data$alpha, method = "spearman")


```


## Alpha vs. Reasoning 

```{r}
data <- full_join(data, models, by=join_by(provider,model))
```

```{r}

grouped_data <- data %>%
  group_by(provider, model, is_reasoner) %>%
  summarize(
    DRI = mean(DRIPostV3_mean, na.rm = TRUE),
    alpha = mean(alpha_policies, na.rm = TRUE)
  )

grouped_data <- grouped_data %>%
  mutate(is_reasoner_clean = ifelse(is.na(is_reasoner), FALSE, is_reasoner))

grouped_data %>%
  ggboxplot(x = "is_reasoner_clean", y = "alpha", add = "jitter")  

labels <- grouped_data[which(grouped_data$alpha > 0.83), ]$model

grouped_data %>%
  group_by(is_reasoner_clean) %>%
  get_summary_stats(alpha, type = "median_iqr")

res.wilcox <- grouped_data %>%
  as.data.frame() %>%
  wilcox_test(alpha ~ is_reasoner_clean, detailed = TRUE) %>%
  add_significance() %>%
  add_xy_position(x = "is_reasoner_clean")
res.wilcox

grouped_data %>%
  as.data.frame() %>%
  wilcox_effsize(alpha ~ is_reasoner_clean)


n_reasoners <- grouped_data %>% filter(is_reasoner_clean) %>% nrow()
n_non_reasoners <- grouped_data %>% filter(!is_reasoner_clean) %>% nrow()

grouped_data %>%
  ggboxplot(
    x = "is_reasoner_clean",
    y = "alpha",
    color = "is_reasoner_clean",
    add = c("jitter"),
    label = "model",
    repel = TRUE,
    label.select = labels,
  ) +
  # ggboxplot(x = "source_type", y = "DRIPostV3", color="source_type") +
  #rotate_x_text(70) +
  stat_pvalue_manual(res.wilcox, tip.length = 0) +
  labs(
    # title = "Human vs. LLM Post-Deliberation DRI",
    caption = get_test_label(res.wilcox, detailed = TRUE),
    #caption = get_pwc_label(pwc),
  ) +
  xlab("Reasoning capability") +
  ylab("Cronbach's alpha (policies)") +
  stat_summary(
    fun.data = function(x)
      data.frame(y = 0.88, label = paste0("\U00B5 = ", round(mean(
        x
      ), 2))),
    geom = "text"
  ) +
  scale_x_discrete(labels = c("TRUE" = paste0("Reasoner (N = ", n_reasoners, ")"), "FALSE" = paste0("Non-Reasoner (N = ", n_non_reasoners, ")"))) +
  theme(legend.position = "none") -> plot

plot

ggsave(
  paste(OUTPUT_DIR, "plots", "book2.png", sep = "/"),
  plot,
  width = 8,
  height = 4
)

```


```{r}

res.wilcox <- grouped_data %>%
  as.data.frame() %>%
  wilcox_test(DRI ~ is_reasoner_clean, detailed = TRUE) %>%
  add_significance() %>%
  add_xy_position(x = "is_reasoner_clean")
res.wilcox

grouped_data %>%
  group_by(is_reasoner_clean) %>%
  get_summary_stats(DRI, type = "median_iqr")

grouped_data %>%
  as.data.frame() %>%
  wilcox_effsize(DRI ~ is_reasoner_clean)

grouped_data %>%
  ggboxplot(
    x = "is_reasoner_clean",
    y = "DRI",
    color = "is_reasoner_clean",
    add = c("jitter")
  ) +
  # ggboxplot(x = "source_type", y = "DRIPostV3", color="source_type") +
  #rotate_x_text(70) +
  labs(
    # title = "Human vs. LLM Post-Deliberation DRI",
    caption = get_test_label(res.wilcox, detailed = TRUE),
    #caption = get_pwc_label(pwc),
  ) +
  xlab("Reasoning capability") +
  ylab("LLM DRI") +
  stat_summary(
    fun.data = function(x)
      data.frame(y = 0.5, label = paste0("\U00B5 = ", round(mean(
        x
      ), 2))),
    geom = "text"
  ) +
  scale_x_discrete(labels = c("TRUE" = paste0("Reasoner (N = ", n_reasoners, ")"), "FALSE" = paste0("Non-Reasoner (N = ", n_non_reasoners, ")"))) +
  theme(legend.position = "none") -> plot

plot

ggsave(
  paste(OUTPUT_DIR, "plots", "book3.png", sep = "/"),
  plot,
  width = 8,
  height = 4
)

```

```{r, fig.width=8,fig.height=4}

selected_data <- grouped_data %>%
  filter(model %like% "grok-3-mini" |
           model %like% "claude-3-7-sonnet-20250219") %>%
  mutate(family = case_when(
    model %like% "grok-3-mini-beta" ~ "grok-3-mini",
    model %like% "grok-3-mini-fast" ~ "grok-3-mini-fast",
    model %like% "claude-3-7-sonnet" ~ "claude-3-7-sonnet",
    .default = NA
  )) %>%
  mutate(label = case_when(
    model %like% "low" ~ "effort = low",
    model %like% "high" ~ "effort = high",
    .default = family
  ))


selected_data %>%
  ggdotplot(
    x="is_reasoner_clean",
    y="DRI",
    label = "label",
    #label.rectangle = TRUE,
    facet.by = "family",
    fill = "is_reasoner_clean",
    color = "is_reasoner_clean",
    ylim = c(0.11,0.41),
    title = "DRI vs. Reasoning Capability"
  ) +
  xlab("Reasoning Capability") +
  ylab("DRI") +
  scale_x_discrete(labels = c("TRUE" = "Reasoner", "FALSE" = "Non-Reasoner")) +
  theme(legend.position = "none") -> plot

plot

ggsave(
  paste(OUTPUT_DIR, "plots", "book4.png", sep = "/"),
  plot,
  width = 8,
  height = 4
)



```

```{r}

nrow(models)

length(unique(data$survey))

length(unique(data$case))

length(unique(data$model))

nrow(llm_data)

nrow(data) * 5

summary(data$alpha_considerations)
summary(data$alpha_policies)

data %>%
  select(alpha_considerations, alpha_policies) %>%
  get_summary_stats(type = "common")

data %>%
  filter(alpha_considerations <= 0 | alpha_policies <= 0) %>%
    select(model, survey, alpha_considerations, alpha_policies)

data %>%
  filter(alpha_considerations >= 1 | alpha_policies >= 1) %>%
  select(model, survey, alpha_considerations, alpha_policies)

```


