---
title: 'Evaluating LLMs Against Human Collective Wisdom: Can AI Reason Deliberatively?'
author: "Francesco Veri, Gustavo Umbelino"
date: "`r Sys.Date()`"
bibliography: bibliography/refs.bib
link-citations: true
csl: bibliography/apsa.csl
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(readxl)

OUTPUT_DIR <- "analysis"

ALPHA_RESULTS_FILE <- "alpha_results.csv"
AGGREGATION_RESULTS_FILE <- "llm_data_aggregated.csv"

read_output_csv <- function(file_name) {
  read_csv(paste(OUTPUT_DIR, file_name, sep = "/"), show_col_types = FALSE)
}

```


# Read in data
```{r read data}

DRI_IND_DATA <- read_output_csv("LLM_alpha_DRI.csv")


```


# Hypotheses Testing

## H1. DRI scores of LLMs do not significantly differ from those produced by a random generation process.

### Testing assumptions

We employed a one-way ANOVA (or a Kruskal-Wallis test, depending on the results of the exploratory analysis) between subjects to analyze our results. If normality and homogeneity of variance assumptions are met, we will use ANOVA followed by Tukey’s HSD post-hoc test for pairwise comparisons between LLM/version DRI and random DRI. If assumptions are violated, we will use the non-parametric Kruskal-Wallis test, followed by Dunn’s post-hoc test with Bonferroni correction.

The independent variable is be the type of participant (e.g., random, model). The dependent variable is the individual-level DRI score.

```{r H1-assumptions, fig.width=10,fig.height=6}

## get data for analysis
DRIInd.ALL <- DRIInd.LLMs %>%
  mutate(source = model) %>%
  mutate(source_type = "llm") %>%
  select(source, source_type, survey, case, DRIPreV3, dv) %>%
  rename(DRIPostV3 = dv)

DRIInd.ALL <- DRIInd.Rands %>%
  mutate(source = "random") %>%
  mutate(source_type = source) %>%
  select(source, source_type, survey, case, DRIPreV3, DRIPostV3) %>%
  bind_rows(DRIInd.ALL)

DRIInd.ALL <- DRI.Global %>%
  filter(type == "human_only") %>%
  mutate(
    source = "human_only_mean",
    source_type = "human",
    DRIPreV3 = DRI_PRE_V3,
    DRIPostV3 = DRI_POST_V3
  ) %>%
  select(source, source_type, survey, case, DRIPreV3, DRIPostV3) %>%
  bind_rows(DRIInd.ALL)

# clean data
DRIInd.ALL <- DRIInd.ALL %>%
  mutate(
    source = as.factor(source),
    source_type = as.factor(source_type),
    survey = as.factor(survey),
    case = as.factor(case)
  )


write_output_csv(DRIInd.ALL, "DRIInd_all.csv")

df <- DRIInd.ALL %>% filter(source != "human_only_mean")

# 1. Normality assumption: Shapiro-Wilk test for each group
shapiro_results <- df %>%
  group_by(source) %>%
  summarise(shapiro_p = shapiro.test(DRIPostV3)$p.value)

# 2. Homogeneity of variances: Levene's Test
levene_test_result <- leveneTest(DRIPostV3 ~ source, data = df)

# plot distributions
df %>%
  full_join(shapiro_results, by = "source") %>%
  mutate(signif = shapiro_p > 0.05) %>%

  ggplot(aes(x = DRIPostV3, fill = source_type)) +
  geom_histogram(bins = 30,
                 alpha = 0.6,
                 position = "identity") +
  facet_wrap(~ source) +
  labs(title = "Distribution of DRIPostV3 for Each Source Type", 
       x = "DRIPostV3", 
       y = "Frequency") +
  theme_minimal() + theme(legend.position = "none") +
  geom_point(
    data = . %>% filter(signif),
    aes(x = -1, y = 5), color = "red", shape = 8, size = 5)

# Assuming your dataframe is named `df`
df %>% arrange(source_type, source) %>%
ggplot(aes(x = source, y = DRIPostV3, fill = source_type)) +
  geom_boxplot() +
  labs(title = "Distribution of DRIPostV3 for Each Source Type",
       x = "Source",
       y = "DRIPostV3") +
  theme_minimal() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1))


```

### Testing hypothesis

```{r H1-test}

kruskal_test_result <- kruskal.test(DRIPostV3 ~ source, data = df)
kruskal_test_result

```

### Post-hoc tests

```{r H1-posthoc}

if (kruskal_test_result$p.value < 0.05) {
  # Perform Dunn's test if Kruskal-Wallis test is significant
  dunn_test_result <- dunn.test(df$DRIPostV3, df$source, method = "bonferroni", kw = FALSE, table = FALSE)
  
  dunn_test_result <- as.data.frame(dunn_test_result)
  
  random_comp <- dunn_test_result %>%
    filter(grepl("random", comparisons)) %>%
    mutate(P.rounded = round(P.adjusted, 3)) %>%
    mutate(P.formated = ifelse(P.rounded < 0.05, paste0(P.rounded, "*"), P.rounded))
  
  # Print the results of Dunn's test with Bonferroni correction
  random_comp %>%
    arrange(P.adjusted) %>% 
    mutate(model = str_remove(comparisons, " - random")) %>%
    select(model, P.formated) %>%
    knitr::kable(
      caption = "Models compared to random",
      col.names = c("Model", "P-adjusted"),
      digits = 3
    )
  
} else {
  message("Kruskal-Wallis test is not significant; no need for post-hoc testing.")
}

```

